Limitations on the Two-Infall Scenario in the Era of Large Stellar Age Catalogs

### 2025-04-18 src/tex/ms.tex
Updates from Overleaf

% user-defined commands
\newcommand{\vice}{{\tt VICE}\xspace}
\newcommand{\todo}[1]{{\color{red}#1}}
\newcommand{\osuaffil}{Department of Astronomy, The Ohio State University, 140 W. 18th Ave, Columbus OH 43210, USA}
\newcommand{\ccappaffil}{Center for Cosmology and AstroParticle Physics, The Ohio State University, 191 W. Woodruff Ave., Columbus OH 43210, USA}
\newcommand{\aFe}{[$\alpha$/Fe]\xspace}
\newcommand{\mathXH}{{\rm [X/H]}}
\newcommand{\mathOH}{{\rm [O/H]}}
\newcommand{\mathFeH}{{\rm [Fe/H]}}
\newcommand{\mathOFe}{{\rm [O/Fe]}}
\newcommand{\mathaFe}{[\alpha/{\rm Fe}]}
\newcommand{\yZ}[1]{$y/Z_\odot=#1$}
\newcommand{\kpc}{\,{\rm kpc}}

@@

An explanation for the $\alpha$-bimodality has yet to be broadly accepted in the GCE community \todo{(reword)}. An $\alpha$-bimodality is not a universal feature in simulated Milky Way-mass galaxies \todo{(citations; also mention external galaxies?)}, so its presence and characteristics in our own galaxy may provide clues to its unique evolutionary history. GCE models that attempt to solve this problem generally fall into two camps: some explain the $\alpha$-bimodality as a result of secular processes, such as the radial migration of stars and the inside-out growth of the disk \todo{(citations)}; others argue that the two sequences point to a bursty star formation history, perhaps driven by multiple discrete accretion events \todo{(citations)} or a change in the star formation efficiency \citep[e.g.][]{conroy_birth_2022}.

@@

\todo{Outflows.} Most previous studies of the two-infall model do not include mass-loaded outflows. Some hydrodynamic simulations of Galactic fountains ejected by CC SNe have shown that ejected material falls back onto the disk on relatively short timescales \citep{spitoni_galactic_2008,spitoni_effects_2009} and close to the point of origin \citep{melioli_hydrodynamical_2008,melioli_hydrodynamical_2009}, suggesting the effect on GCE could be minimal. However, other simulations of MW-like galaxies with different feedback prescriptions do produce mass-loaded outflows \todo{(citations)}. Empirically, mass-loaded outflows have been observed in nearby starburst galaxies \todo{(citations)} and dwarf galaxies \todo{(citations)} but not MW-like systems due to current detection limits \todo{(citations)}. We adopt an {\it ad hoc} prescription for outflows as a function of radius in order to balance different yield sets and reproduce the radial metallicity gradient.

\todo{Migration.} The two-infall model attempts to reproduce the full distribution of stellar abundances in the Solar neighborhood through a single, continuous evolutionary track. \citet{sellwood_radial_2002} first showed that transient spiral perturbations can induce large changes in the guiding radius of a star without kinematic heating. It is now understood that the stars that make up the Solar neighborhood are drawn from a wide range of birth radii \citep[e.g.,][]{schonrich_chemical_2009,frankel_measuring_2018,lehmann_probing_2024}. Some studies have attempted to derive stellar birth radii \citep[e.g.,][]{ratcliffe_unveiling_2023,lu_there_2024}, though such an endeavor requires assumptions about the evolution of the Milky Way's radial metallicity gradient. While the strength and speed of radial migration in the disk are not precisely measured, it is clear that a single chemical evolution track need not explain the entirety of the local observed abundance distribution.

\todo{Summary of secular evolution models which claim alpha-bimodality.} \citet{schonrich_chemical_2009}. \citet{kubryk_evolution_2015}. \citet{sharma_chemical_2021}, \citet{chen_chemical_2023}. \citet{prantzos_origin_2023}.

\todo{Introduce equilibrium model.}

@@

\subsection{The Scale of Yields \& Outflows}
\label{sec:multizone-yields}

@@

We first examine the differences between multi-zone models which assume different yield and outflow scales. Figure \ref{fig:abundance-evolution-yields} shows stellar age--abundance relations produced by models with \yZ{1} and \yZ{2} with fiducial parameters (Table \ref{tab:multizone-parameters}). The dilution effect discussed in Section \ref{sec:yields} is clearly seen here. The \yZ{1} model (column a) shows two clear discrepancies with the \citet{leung_variational_2023} age--abundance relation: a major $\sim0.5$ dex dilution at a lookback time of $\sim9\,{\rm Gyr}$ near where the data show a maximum in [O/H], and non-zero abundance evolution at late times where the data show very little abundance evolution. The evolution of [Fe/H] is similar, but the approach to the final metallicity is slower because of the additional delay imposed on Fe production from SNe Ia. The \yZ{2} yield set (column b) mitigates both of these issues by shortening the time it takes the ISM metallicity to rebound post-$t_{\rm max}$, producing a much flatter abundance curve at late times. However, model (b) produces a poorer fit to the age--[O/Fe] relation: the decline in [O/Fe] over the thin disk epoch is steeper than the data, especially for ages $\sim4-8\,{\rm Gyr}$.

\subsection{Local Abundance Evolution}
\label{sec:age-abundance}

@@

We attempt to mitigate these discrepancies in the age--metallicity relation for the empirical yield scale. 
% Figure \ref{fig:abundance-evolution-params} shows the effect of varying several parameters on the \yZ{1} model. In column (b), we approximately double the strength of radial migration to $\sigma_{\rm RM8}=5.0\kpc$.
For instance, the observed rise in the median metallicity of stars with ages of $\sim4-10\,{\rm Gyr}$ is thought to be due to radial migration, as those stars are more likely to have migrated from the dense inner metal-rich regions of the Galaxy during that time. Although our fiducial model does include a prescription for radial migration, the majority of stars in that age range in Figure \ref{fig:abundance-evolution-yields} still have sub-Solar abundances. Therefore, column (b) of Figure \ref{fig:abundance-evolution-params} presents a model with $y/Z_\odot=1$ and a stronger migration prescription of $\sigma_{\rm RM8}=5\kpc$. As a result, the stars which make up the present-day Solar neighborhood are drawn from a wider range of birth $R_{\rm gal}$, producing a broader abundance distribution for any given age. However, even though this prescription is much stronger than the estimates of, e.g., \citet{frankel_measuring_2018}, the model still significantly under-predicts the metallicity of $\sim4-9\,{\rm Gyr}$ old stars.

Next, as in Section \ref{sec:abundance-distributions}, we investigate a model where the infalling gas is enriched to a metallicity $\mathOH=\mathFeH={\rm [X/H]}_{\rm CGM}$ before accreting onto the disk. Column (c) of Figure \ref{fig:abundance-evolution-params} shows results for the case where ${\rm [X/H]}_{\rm CGM}-0.5$, the highest metallicity allowed by the local low-$\alpha$ population. Pre-enriched infall at this level mitigates but does not completely solve the two discrepancies. The dilution effect of the second infall is reduced to the $\sim0.3$-dex level as the gas which replenishes the Galaxy's reservoir is no longer pristine; however, the width of the stellar abundance distribution at any given age is also reduced, because the enriched gas accretion imposes a lower limit on the metallicity of the outermost regions, from which the stars of the low-metallicity tail in the Solar neighborhood are drawn. The late-time gas abundance evolution is similar to the fiducial model, but it ends at slightly super-Solar metallicity --- an effect which can be compensated by a slightly increased value of $\eta$. This model also narrows the [O/Fe] distribution of mono-age populations (almost all the model stars fall within the $1\sigma$ band of the data), which could be compensated for by stronger radial migration.

Finally, we explore a model where the local thick-to-thin disk surface density ratio is $\sim4$ times larger, $f_\Sigma(R_\odot)=0.48$. This is higher than most of the constraints from population counts or GCE models (see Section \ref{sec:sfh}). Column (d) of Figure \ref{fig:abundance-evolution-params} shows that requiring a more massive thick disk can reduce the dilution and recent evolution of the ISM, similar to the pre-enriched infall, because more of the gas disk is built up during the first infall phase. The model produces the best agreement with the observed age--[Fe/H] relation (second row). However, agreement with the observed age--[O/Fe] relation is poor, with the model predicting less evolution in [O/Fe] over the past $\sim9$ Gyr than observed by $\approx0.1$ dex.

@@

\subsection{Global [O/Fe] Distributions: Two or Three Peaks?}
\label{sec:disk-abundances}

@@

\subsection{Age Patterns}

@@

@@

@@

>>>>

% user-defined commands
\newcommand{\vice}{{\tt VICE}\xspace}
\newcommand{\todo}[1]{{\color{red}#1}}
\newcommand{\osuaffil}{Department of Astronomy, The Ohio State University, 140 W. 18th Ave, Columbus OH 43210, USA}
\newcommand{\ccappaffil}{Center for Cosmology and AstroParticle Physics, The Ohio State University, 191 W. Woodruff Ave., Columbus OH 43210, USA}
\newcommand{\aFe}{[$\alpha$/Fe]\xspace}
\newcommand{\mathXH}{{\rm [X/H]}}
\newcommand{\mathOH}{{\rm [O/H]}}
\newcommand{\mathFeH}{{\rm [Fe/H]}}
\newcommand{\mathOFe}{{\rm [O/Fe]}}
\newcommand{\mathaFe}{[\alpha/{\rm Fe}]}
\newcommand{\yZ}[1]{$y/Z_\odot=#1$}
\newcommand{\kpc}{\,{\rm kpc}}
\newcommand{\Gyr}{\,{\rm Gyr}}

@@

An explanation for the $\alpha$-bimodality has yet to be broadly accepted in the GCE community \todo{(reword)}. An $\alpha$-bimodality is not a universal feature in simulated Milky Way-mass galaxies \todo{(citations; also mention external galaxies?)}, so its presence and characteristics in our own galaxy may provide clues to its unique evolutionary history. GCE models that attempt to solve this problem generally fall into two camps: some explain the $\alpha$-bimodality as a result of secular processes, such as the radial migration of stars and the inside-out growth of the disk \todo{(citations)}; others argue that the two sequences point to a bursty star formation history, perhaps driven by multiple discrete accretion events \citep[e.g.,][]{chiappini_chemical_1997,mackereth_origin_2018,spitoni_beyond_2023} or a change in the star formation efficiency \citep[e.g.][]{conroy_birth_2022}.

@@

% \todo{Outflows.} 
Most previous studies of the two-infall model do not include mass-loaded outflows. Some hydrodynamic simulations of Galactic fountains ejected by CC SNe have shown that ejected material falls back onto the disk on relatively short timescales \citep{spitoni_galactic_2008,spitoni_effects_2009} and close to the point of origin \citep{melioli_hydrodynamical_2008,melioli_hydrodynamical_2009}, suggesting the effect on GCE could be minimal. However, other simulations of MW-like galaxies with different feedback prescriptions do produce mass-loaded outflows \todo{(citations)}. Empirically, mass-loaded outflows have been observed in nearby starburst galaxies \todo{(citations)} and dwarf galaxies \todo{(citations)} but not MW-like systems due to current detection limits \todo{(citations)}. We adopt an {\it ad hoc} prescription for outflows as a function of radius in order to balance different yield sets and reproduce the radial metallicity gradient. \todo{Trim down, move some to methods.}

% \todo{Migration.} 
The two-infall model attempts to reproduce the full distribution of stellar abundances in the Solar neighborhood through a single, continuous evolutionary track. \citet{sellwood_radial_2002} first showed that transient spiral perturbations can induce large changes in the guiding radius of a star without kinematic heating. It is now understood that the stars that make up the Solar neighborhood are drawn from a wide range of birth radii \citep[e.g.,][]{schonrich_chemical_2009,frankel_measuring_2018,lehmann_probing_2024}. Some studies have attempted to derive stellar birth radii \citep[e.g.,][]{ratcliffe_unveiling_2023,lu_there_2024}, though such an endeavor requires assumptions about the evolution of the Milky Way's radial metallicity gradient. While the strength and speed of radial migration in the disk are not precisely measured, it is clear that a single chemical evolution track need not explain the entirety of the local observed abundance distribution.

\todo{Summary of secular evolution models which claim alpha-bimodality.} By incorporating stellar ``churning'' (changes in guiding-center radius) and ``blurring'' (changes in orbital eccentricity) as well as radial gas flows in a multi-ring model, \citet{schonrich_chemical_2009} successfully produced distinct high- and low-$\alpha$ sequences through secular processes alone; however, they failed to reproduce the overlap in metallicity between the two that would be a key finding of later surveys \citep[e.g.,][]{bensby_exploring_2014}. \citet{kubryk_evolution_2015} produce two parallel sequences in \aFe vs [Fe/H] by making a simple age cut to the results of their multi-zone GCE model, though because they do not present stellar \aFe distributions it is hard to determine whether their distributions are truly bimodal. In the model of \citet{sharma_chemical_2021}, a sharp transition in stellar \aFe for $8-12\Gyr$ old stars plus radial migration are the key ingredients for predicting a MW-like bimodality; however, their approach to the chemical evolution is not {\it ab initio}, meaning that instead of integrating the enrichment equations for their adopted star formation history, they simply assume a functional form for the abundance evolution over time \todo{(rewrite)}. \todo{Sharma basically says that it's possible to have a continuous star formation history and reproduce the $\alpha$-bimodality.} \citet{chen_chemical_2023} confirmed the findings of \citet{sharma_chemical_2021} in an {\it ab initio} multi-zone GCE model (with a two-phase ISM and radial gas flows but no outflows) and were able to produce a bimodal \aFe distribution; however, some discrepancies remain, including that the metallicity of the high-$\alpha$ sequence is too low and not extended enough. \citet{chen_chemical_2023} found that the abundance gap is caused by the simultaneous decline in SFR and \aFe. \citet{prantzos_origin_2023}.

% \todo{Introduce equilibrium model.} 
As stellar age estimation techniques have improved over recent years, large catalogs have become available with ages for hundreds of thousands or even millions of stars from a wide swath of the Galaxy. In a challenge to the traditional view of GCE, which expects the ISM metallicity to continually increase over time, \citet{johnson_milky_2024} examined the age--metallicity relation at different radii from the astroNN catalog \citep{mackereth_dynamical_2019} and found that the mode of the metallicity distribution at a given radius is nearly independent from age over the past $\sim9\Gyr$. They propose an ``equilibrium scenario'' in which the local metallicity is driven by the ratio of star formation to accretion at a given radius, which evolves to a constant over $\sim\Gyr$ timescales. Whether the equilibrium metallicity is driven by outflows, as proposed by \citet{johnson_milky_2024}, or by other factors such as a radial gas flow, the current data suggest that the gas abundance across the Galaxy has evolved very little over most of the thin disk lifetime.

@@

\subsection{Dilution \& Approach to Equilibrium}
\label{sec:age-abundance}

@@

The dilution effect discussed in Section \ref{sec:yields} is clearly seen in the multi-zone model results. We first examine the differences between multi-zone models which assume different yield and outflow scales. Figure \ref{fig:abundance-evolution-yields} shows stellar age--abundance relations produced by models with \yZ{1} and \yZ{2} with fiducial parameters (Table \ref{tab:multizone-parameters}). The \yZ{1} model (column a) shows two clear discrepancies with the \citet{leung_variational_2023} age--abundance relation: a major $\sim0.5$ dex dilution at a lookback time of $\sim9\,{\rm Gyr}$ near where the data show a maximum in [O/H], and non-zero abundance evolution at late times where the data show very little abundance evolution. The evolution of [Fe/H] is similar, but the approach to the final metallicity is slower because of the additional delay imposed on Fe production from SNe Ia. The \yZ{2} yield set (column b) mitigates both of these issues by shortening the time it takes the ISM metallicity to rebound post-$t_{\rm max}$, producing a much flatter abundance curve at late times. However, model (b) produces a poorer fit to the age--[O/Fe] relation: the decline in [O/Fe] over the thin disk epoch is steeper than the data, especially for ages $\sim4-8\,{\rm Gyr}$.

@@

We next attempt to mitigate the dilution and late-time evolution problems for the empirical (\yZ{1}) yield scale. 
Figure \ref{fig:abundance-evolution-params} shows the effect of varying several parameters for the \yZ{1} model: (b) the strength of radial migration $\sigma_{\rm RM8}$, (c) the metallicity of the infalling gas $\mathXH_{\rm CGM}$, and (d) the local thick-to-thin disk density ratio $f_\Sigma(R_\odot)$.

The observed rise in the median metallicity of stars with ages of $\sim4-10\,{\rm Gyr}$ is thought to be due to radial migration, as those stars are more likely to have migrated from the dense inner metal-rich regions of the Galaxy during that time. Although our fiducial model does include a prescription for radial migration, the majority of stars in that age range in Figure \ref{fig:abundance-evolution-yields} still have sub-Solar abundances. Therefore, column (b) of Figure \ref{fig:abundance-evolution-params} presents a model with $y/Z_\odot=1$ and a stronger migration prescription of $\sigma_{\rm RM8}=5\kpc$. As a result, the stars which make up the present-day Solar neighborhood are drawn from a wider range of birth $R_{\rm gal}$, producing a broader abundance distribution for any given age. However, even though this prescription is much stronger than the estimates of, e.g., \citet{frankel_measuring_2018}, the model still significantly under-predicts the metallicity of $\sim4-9\,{\rm Gyr}$ old stars.

Next, we investigate a model where the infalling gas is enriched to a metallicity $\mathOH=\mathFeH={\rm [X/H]}_{\rm CGM}$ before accreting onto the disk. Column (c) of Figure \ref{fig:abundance-evolution-params} shows results for the case where ${\rm [X/H]}_{\rm CGM}=-0.5$, the highest metallicity allowed by the local low-$\alpha$ population \todo{(reword}). Pre-enriched infall at this level mitigates but does not completely solve the two discrepancies. The dilution effect of the second infall is reduced to the $\sim0.3$-dex level as the gas which replenishes the Galaxy's reservoir is no longer pristine; however, the width of the stellar abundance distribution at any given age is also reduced, because the enriched gas accretion imposes a lower limit on the metallicity of the outermost regions, from which the stars of the low-metallicity tail in the Solar neighborhood are drawn. The late-time gas abundance evolution is similar to the fiducial model, but it ends at slightly super-Solar metallicity---an effect which can be compensated by a slightly increased value of $\eta$. This model also narrows the [O/Fe] distribution of mono-age populations (almost all the model stars fall within the $1\sigma$ band of the data), which could be compensated for by stronger radial migration.

Finally, we explore a model where the local thick-to-thin disk surface density ratio is $\sim4$ times larger that the fiducial value, $f_\Sigma(R_\odot)=0.5$. This is higher than most of the constraints from population counts or GCE models (see Section \ref{sec:sfh}). Column (d) of Figure \ref{fig:abundance-evolution-params} shows that requiring a more massive thick disk can reduce the dilution and recent evolution of the ISM, similar to the pre-enriched infall, because more of the gas disk is built up during the first infall phase. The model produces the best agreement with the observed age--[Fe/H] relation (second row). However, agreement with the observed age--[O/Fe] relation is poor, with the model predicting less evolution in [O/Fe] over the past $\sim9$ Gyr than observed by $\approx0.1$ dex.

@@

\subsection{Global Abundance Distributions}
\label{sec:disk-abundances}

\subsubsection{The [O/Fe] Distribution: Two or Three Peaks?}

@@

\subsubsection{The Best Model}
\label{sec:ofe-feh-best}

@@

\subsection{Local Age Patterns}

@@

The two-infall model makes a prediction about the local stellar age distribution that is fundamental to its construction: that the most metal-rich stars in any region of the Galaxy come from the metal-rich tail of the first infall sequence, and are therefore older than all of the thin disk stars. This prediction can be seen in any of the panels in Figure \ref{fig:ofe-feh-best}. In Figure \ref{fig:lmr-ages}, there is a sharp divide in the median age of local metal-rich (LMR; $\mathFeH\gtrsim+0.1$) stars when moving from the thick disk ($\tau\ge9\Gyr$) to the thin disk ($\tau\lesssim5\Gyr$).

@@

\subsection{Third Accretion Episode}
